Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-09 manuscript/manuscript.tex
purged the neutral

 Rapid phenotypic adaptation is often observed in natural populations and
 selection experiments.  However, detecting the genome-wide impact of this
 selection is difficult, since adaptation often proceeds from standing
 variation and selection on highly polygenic traits, both of which may leave
 faint genomic signals indistinguishable from a noisy background of genetic
 drift.  One promising signal comes from the genome-wide covariance between
 allele frequency changes observable from temporal genomic data, e.g.
 evolve-and-resequence studies. These temporal covariances reflect how the
 change in neutral allele frequency at one timepoint is predictive of the
 changes at later timepoints when there is heritable fitness variation in the
 population, as neutral alleles can remain associated with selected alleles
 over time. Since genetic drift does not lead to temporal covariance, we can
 use these covariances to estimate what fraction of the variation in allele
 frequency change through time is driven by linked selection. Here, we
 reanalyze \vb{three selection experiments}, \vbout{two \emph{Drosophila
   simulans} evolve-and-resequence studies, and one artificial selection
 experiment in mice}, to quantify the effects of linked selection over short
 timescales using covariance among time-points and across replicates. We
 estimate that at least 17\% to 37\% of allele frequency change is driven by
 selection in these experiments.  Against this background of positive
 genome-wide temporal covariances we also identify signals of negative temporal
 covariance corresponding to reversals in the direction of selection for a
 reasonable proportion of loci over the time course of a selection experiment.
 Overall, we find that in the three studies we analyzed, linked selection has a
 large impact on short-term allele frequency dynamics that is readily
 distinguishable from genetic drift.

@@

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid,
selection-driven genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact of selection on genome-wide variation
over short timescales in sexual populations remarkably difficult
\vb{\parencite{Latta1998-me, Pritchard2010-tk,Kemper2014-bx}}. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across \vb{numerous} loci. This can lead to subtle allele frequency shifts on
standing variation that are difficult to distinguish from background levels of
genetic drift and sampling variance. \vb{Increasingly}, genomic experimental
evolution studies with multiple timepoints, and in some cases multiple
replicate populations, are being used to detect large effect selected loci
\parencite{Turner2011-sx,Turner2012-bm} and differentiate modes of selection
\parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In addition these
temporal-genomic studies have begun in wild populations, some with the goal of
finding variants that exhibit frequency changes consistent with fluctuating
selection \parencite{Bergland2014-ij,Machado2018-cs}. In a previous paper, we
proposed that one useful signal for understanding the genome-wide impact of
polygenic linked selection detectable from genomic studies with multiple
timepoints is the temporal autocovariance (\vb{i.e. covariance between two
timepoints}) in allele frequency changes \parencite{Buffalo2019-io}.  These
covariances \vb{can be estimated} from temporal genomic data and are created
when the loci that underly heritable fitness variation perturb the frequencies
of linked neutral alleles; in contrast, when genetic drift acts alone in a
closed population, these covariances are zero in expectation.  Mathematically,
temporal covariances are useful because it is natural to decompose the total
variance in allele frequency change across a set of time intervals into the
variances and covariances in allele frequency change among time intervals.
Furthermore, biologically, these covariances reflect the extent to which
neutral allele frequency changes in one generation predict changes in another
due to a shared selection pressures and associations to selected loci.

@@

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals (Figure \ref{fig:figure-1} A), as expected when directional
selection affects linked variants' frequency trajectories until ultimately
linkage disequilibrium and the additive genetic variance for fitness associated
with neutral alleles decays (\vb{which could occur as a population reaches
a new optimum}) \parencite{Buffalo2019-io}. The temporal covariances per
replicate are noisier but this general pattern holds; see Supplementary Figure
\ref{suppfig:barghi-cov-panels}.  \textcite{Barghi2019-qy}'s design means that
the covariances we see in adjacent time intervals are on average ten
generations apart, and given the temporal decay in covariance we see, the
covariances on shorter time-scales (e.g. if adjacent generations had been
sequenced) may well be higher yet (see Supplementary Material Section
\ref{supp:barghi-covs} for more details).

@@

The replicate design of \textcite{Barghi2019-qy} allows us to quantify another
covariance: the covariance in allele frequency change between replicate
populations experiencing convergent selection pressures. These
between-replicate covariances are created in the same way as temporal
covariances: neutral alleles linked to a particular fitness background are
expected to have allele frequency changes in the same direction if the
selection pressures are similar. Intuitively, where temporal covariances
reflect that neutral alleles associated with heritable fitness backgrounds are
predictive of frequency changes between generations, replicate covariances
reflect that heritable fitness backgrounds common to each replicate predict
(under the same selection pressures) frequency changes between replicates. We
measure this through a statistic similar to a correlation, which we call the
convergent correlation: the ratio of average between-replicate covariance
across all pairs to the average standard deviation across all pairs of
replicates,

>>>>

 Rapid phenotypic adaptation is often observed in natural populations and
 selection experiments.  However, detecting the genome-wide impact of this
 selection is difficult, since adaptation often proceeds from standing
 variation and selection on highly polygenic traits, both of which may leave
 faint genomic signals indistinguishable from a noisy background of genetic
 drift.  One promising signal comes from the genome-wide covariance between
 allele frequency changes observable from temporal genomic data, e.g.
 evolve-and-resequence studies. These temporal covariances reflect how the
 change \gc{in allele} frequency at one timepoint is predictive of the
 changes at later timepoints when there is heritable fitness variation in the
 population, \gc{as alleles} can remain associated with selected alleles
 over time \gc{and so be indirectly selected for}. Since genetic drift does not lead to temporal covariance, we can
 use these covariances to estimate what fraction of the variation in allele
 frequency change through time is driven by linked selection. Here, we
 reanalyze \vb{three selection experiments}, \vbout{two \emph{Drosophila
   simulans} evolve-and-resequence studies, and one artificial selection
 experiment in mice}, to quantify the effects of linked selection over short
 timescales using covariance among time-points and across replicates. We
 estimate that at least 17\% to 37\% of allele frequency change is driven by
 selection in these experiments.  Against this background of positive
 genome-wide temporal covariances we also identify signals of negative temporal
 covariance corresponding to reversals in the direction of selection for a
 reasonable proportion of loci over the time course of a selection experiment.
 Overall, we find that in the three studies we analyzed, linked selection has a
 large impact on short-term allele frequency dynamics that is readily
 distinguishable from genetic drift.

@@

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid,
selection-driven genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact of selection on genome-wide variation
over short timescales in sexual populations remarkably difficult
\vb{\parencite{Latta1998-me, Pritchard2010-tk,Kemper2014-bx}}. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across \vb{numerous} loci. This can lead to subtle allele frequency shifts on
standing variation that are difficult to distinguish from background levels of
genetic drift and sampling variance. \vb{Increasingly}, genomic experimental
evolution studies with multiple timepoints, and in some cases multiple
replicate populations, are being used to detect large effect selected loci
\parencite{Turner2011-sx,Turner2012-bm} and differentiate modes of selection
\parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In addition these
temporal-genomic studies have begun in wild populations, some with the goal of
finding variants that exhibit frequency changes consistent with fluctuating
selection \parencite{Bergland2014-ij,Machado2018-cs}. In a previous paper, we
proposed that one useful signal for understanding the genome-wide impact of
polygenic linked selection detectable from genomic studies with multiple
timepoints is the temporal autocovariance (\vb{i.e. covariance between two
timepoints}) in allele frequency changes \parencite{Buffalo2019-io}.  These
covariances \vb{can be estimated} from temporal genomic data and are created
when the loci that underly heritable fitness variation perturb the frequencies
of linked \gc{alleles}; in contrast, when genetic drift acts alone in a
closed population, these covariances are zero in expectation \gc{for
  neutral alleles}.  Mathematically,
temporal covariances are useful because it is natural to decompose the total
variance in allele frequency change across a set of time intervals into the
variances and covariances in allele frequency change among time intervals.
Furthermore, biologically, these covariances reflect the extent to which
\gc{allele} frequency changes in one generation predict changes in another
due to a shared selection pressures and associations to selected loci.

@@

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals (Figure \ref{fig:figure-1} A), as expected when directional
selection affects linked variants' frequency trajectories until ultimately
linkage disequilibrium and the \gc{associated} additive genetic
variance for fitness decays (\vb{which could occur as a population reaches
a new optimum}) \parencite{Buffalo2019-io}. The temporal covariances per
replicate are noisier but this general pattern holds; see Supplementary Figure
\ref{suppfig:barghi-cov-panels}.  \textcite{Barghi2019-qy}'s design means that
the covariances we see in adjacent time intervals are on average ten
generations apart, and given the temporal decay in covariance we see, the
covariances on shorter time-scales (e.g. if adjacent generations had been
sequenced) may well be higher yet (see Supplementary Material Section
\ref{supp:barghi-covs} for more details).

@@

The replicate design of \textcite{Barghi2019-qy} allows us to quantify another
covariance: the covariance in allele frequency change between replicate
populations experiencing convergent selection pressures. These
between-replicate covariances are created in the same way as temporal
covariances: \gc{alleles} linked to a particular fitness background are
expected to have allele frequency changes in the same direction if the
selection pressures are similar. Intuitively, where temporal covariances
reflect that \gc{alleles} associated with heritable fitness backgrounds are
predictive of frequency changes between generations, replicate covariances
reflect that heritable fitness backgrounds common to each replicate predict
(under the same selection pressures) frequency changes between replicates. We
measure this through a statistic similar to a correlation, which we call the
convergent correlation: the ratio of average between-replicate covariance
across all pairs to the average standard deviation across all pairs of
replicates,
